---
title: "3 Questions for Checkpoint 3"
author: "Ashley Martinez, Drew Collopy, Hannah Stackpole"
editor: visual
format: 
  html:  
    embed-resources: true
    code-tools: true
    toc: true
---

# Setup

```{r}
#load packages here
library(tidyverse)
```

# About the Data

The data for this lab comes from The Stanford Education Data Archive (SEDA), which is maintained by the Educational Opportunity Project (EOP) at Stanford University (https:\\edopportunity.org). There is a plethora of nationwide data on academic achievement and school demographics/inequality.

We will focus on a "long" dataset that contains math and Reading/English Language Arts (RLA) performance for administrative school districts in the U.S. for the 2008-09 through 2018-19 and 2021-22 through 2024-25 school years.

While the construction of the data uses a complex model, it is important to understand that these are not raw test scores. Rather, they are cohort-standardized scores. This means that the values can be interpreted as the number of standard deviations above or below the national average scores of four reference groups (students who were in fourth grade in the years 2009, 2011, 2013, and 2015).

Practically speaking, this means that the estimates are comparable across time and across places, but not across grades or subjects. Please keep this in mind when designing graphs/analyses. If you would like to learn more about our data, you can check out the following resources:

-   [Technical Documentation](https://stacks.stanford.edu/file/hm970gr1371/SEDA_documentation_v2025.1.pdf#page=4.36)
-   [SEDA site (information about the project and downloadable codebooks with variable descriptions)](https://edopportunity.org/trends/data/downloads/#testscore-2025)

*Note: the relevant codebooks are `seda_codebook_admindist_2025.1` and `seda_codebook_cov_admindist_2025.1`. You will need to navigate to the sheets that match the file names for our specific datasets.*

Below are descriptions of the test data variables we will focus on, but you are encouraged to check out the codebook if you wish to explore other variables.

| Variable        | Description                                     |
|-----------------|-------------------------------------------------|
| `sedaadmin`     | SEDA Admin ID                                   |
| `sedaadminname` | Administrative District Name                    |
| `subject`       | Tested Subject                                  |
| `grade`         | Tested Grade                                    |
| `year`          | Spring of Tested Year                           |
| `fips`          | State FIPS Code                                 |
| `stateabb`      | State Abbreviation                              |
| `cs_mn_all`     | Admin. Dist. Subject-Grade-Year Average         |
| `tot_asmt_all`  | Total Number Tested in the Subject-Grade-Year   |
| `cs_mn_fem`     | Admin. Dist. Subject-Grade-Year Average Female  |
| `tot_asmt_fem`  | Number Females Tested in the Subject-Grade-Year |
| `cs_mn_mal`     | Admin. Dist. Subject-Grade-Year Average Male    |
| `tot_asmt_mal`  | Number Males Tested in the Subject-Grade-Year   |

The covariate data contains a variety of data on city size, school locale, school demographics, student poverty rates (measured by free/reduced school lunch status), household income and unemployment data, and school diversity/"segregation" measures.

# Reading in the Data

```{r}
test_data <- read_csv("seda_admindist_long_cs_2025.1.csv")
cov_data <- read_csv("seda_cov_admindist_long_2025.1.csv")
```

# Cleaning and Understanding the Data

**Question 1: Cleaning the data** Before we begin our analysis, we need to clean and join our two datasets. Complete the following steps: 1. For test_data, filter it to only include states that have the abbreviation CA for California 2. Then select the columns shown in the "About the Data" above 3. Rename the columns to something that is easier to read 4. Now join the two datasets

```{r}
test_data_clean <- test_data |> 
  filter(stateabb == "CA") |> 
  select(sedaadmin, sedaadminname, subject, grade, year, fips, stateabb, cs_mn_all, tot_asmt_all, cs_mn_fem, tot_asmt_fem, cs_mn_mal, tot_asmt_mal) |>
  rename()
```

Question 2: Understanding the data summary table
- function called summary_table() 
- use test_data dataset which was orginally seda admindist long and use variables subject, grade, and cs_mn_all 
-Subject= Tested subject(Math or RLA), grade= Tested Grade , cs_mn_all = Average achievement score for all students
- group by subject and grade level 
- find the average achievement score using cs_mn_all 
- summarize the number of observations within each group
- Determine any missing values
- The summary table will then be returned. This will enable us to comprehend grade-level and subject patterns. 

**Question 3: ggplot Function** Next, we will write a function that operates on the original test scores dataset, with the following requirements:

-   operates on the original `test_data` that is not limited to CA data
-   takes **three** arguments: a grade, a variable of interest, and a list of state abbreviations
-   uses ggplot2 to create **two side-by-side graphs**. The graphs should be RLA and math, and depict the variable of interest average\* (remember our embrace and walrus operators!) **for each state for the specified grade**
-   contains input validation to ensure the grade is numeric and there are **no more than three states** specified
-   applies a **ggplot theme** of your choosing (ggthemes or ggplot2 themes)

\*The average will not be the best state-wide measure of the variable as it is the average of the district data, not weighted by district size. We will ignore this limitation for the sake of function-writing practice!

*Note: We will provide an example output graph to make sure students understand the format requireed*
