---
title: "3 Questions for Checkpoint 3"
format: html
editor: visual
---

```{r}
library(tidyverse)
test_data <- read_csv("seda_admindist_long_cs_2025.1.csv")
```

```{r}
cov_data <- read_csv("seda_cov_admindist_long_2025.1.csv")
```

Question 1: Cleaning the data

```{r}
test_data_clean <- test_data |> 
  filter(stateabb == "CA") |> 
  select(sedaadmin, sedaadminname, subject, grade, year, fips, stateabb, cs_mn_all, tot_asmt_all, cs_mn_fem, tot_asmt_fem, cs_mn_mal, tot_asmt_mal) |>
  rename()
```

select columns for second data set join data sets

Question 2: Understanding the data summary table

**Question 3:** Function
Next, we will write a function that operates on the original test scores dataset, with the following requirements:
- operates on the original `test_data` that is not limited to CA data 
- takes **three** arguments: a grade, a variable of interest, and a list of state abbreviations
- uses ggplot2 to create **two side-by-side graphs**. The graphs should be RLA and math, and depict the variable of interest average* (remember our embrace and walrus operators!) **for each state for the specified grade** 
- contains input validation to ensure the grade is numeric and there are **no more than three states** specified 
- applies a **ggplot theme** of your choosing (ggthemes or ggplot2 themes)

*The average will not be the best state-wide measure of the variable as it is the average of the district data, not weighted by district size. We will ignore this limitation for the sake of function-writing practice!

*Note: We will provide an example output graph to make sure students understand the format requireed*
