---
title: "3 Questions for Checkpoint 3"
author: "Ashley Martinez, Drew Collopy, Hannah Stackpole"
editor: visual
format: 
  html:  
    embed-resources: true
    code-tools: true
    toc: true
---

# Setup

```{r}
#load packages here
library(tidyverse)
```

# About the Data

| Variable        | Description                                     |
|-----------------|-------------------------------------------------|
| `sedaadmin`     | SEDA Admin ID                                   |
| `sedaadminname` | Administrative District Name                    |
| `subject`       | Tested Subject                                  |
| `grade`         | Tested Grade                                    |
| `year`          | Spring of Tested Year                           |
| `fips`          | State FIPS Code                                 |
| `stateabb`      | State Abbreviation                              |
| `cs_mn_all`     | Admin. Dist. Subject-Grade-Year Average         |
| `tot_asmt_all`  | Total Number Tested in the Subject-Grade-Year   |
| `cs_mn_fem`     | Admin. Dist. Subject-Grade-Year Average Female  |
| `tot_asmt_fem`  | Number Females Tested in the Subject-Grade-Year |
| `cs_mn_mal`     | Admin. Dist. Subject-Grade-Year Average Male    |
| `tot_asmt_mal`  | Number Males Tested in the Subject-Grade-Year   |

# Reading in the Data

```{r}
test_data <- read_csv("seda_admindist_long_cs_2025.1.csv")
cov_data <- read_csv("seda_cov_admindist_long_2025.1.csv")
```

# Cleaning and Understanding the Data

**Question 1: Cleaning the data**
Before we begin our analysis, we need to clean and join our two datasets. Complete the following steps:
1. For test_data, filter it to only include states that have the abbreviation CA for California
2. Then select the columns shown in the "About the Data" above
3. Rename the columns to something that is easier to read
4. Now join the two datasets

```{r}
test_data_clean <- test_data |> 
  filter(stateabb == "CA") |> 
  select(sedaadmin, sedaadminname, subject, grade, year, fips, stateabb, cs_mn_all, tot_asmt_all, cs_mn_fem, tot_asmt_fem, cs_mn_mal, tot_asmt_mal) |>
  rename()
```


Question 2: Understanding the data summary table

# Functions

**Question 3: ggplot Function** Next, we will write a function that operates on the original test scores dataset, with the following requirements:

-   operates on the original `test_data` that is not limited to CA data
-   takes **three** arguments: a grade, a variable of interest, and a list of state abbreviations
-   uses ggplot2 to create **two side-by-side graphs**. The graphs should be RLA and math, and depict the variable of interest average\* (remember our embrace and walrus operators!) **for each state for the specified grade**
-   contains input validation to ensure the grade is numeric and there are **no more than three states** specified
-   applies a **ggplot theme** of your choosing (ggthemes or ggplot2 themes)

\*The average will not be the best state-wide measure of the variable as it is the average of the district data, not weighted by district size. We will ignore this limitation for the sake of function-writing practice!

*Note: We will provide an example output graph to make sure students understand the format requireed*
